---
title: "stratification-model"
author: "Sam Bashevkin"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = 'docs',
      knit_root_dir = "../",
      envir = globalenv()
    )
    })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

packages
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(GGally)
library(mgcv)
library(itsadug)
library(downloadthis)
```

read data
```{r}
test_train_cutoff<-as_date("2022-09-29")

d<-read_csv("data/X2_ModelInputs_v2.csv")%>%
  filter(if_any(everything(), ~ !is.na(.x)))%>% # Remove empty rows
  rename(NDOI=`NDOI (cfs)`, Weak_ebb_flood=`Weak Ebb-Flood (m)`, Stratification=`Stratification (PSU)`,
         X2_surface=`X2_TopSalinity (km)`, X2_bottom=`X2_BottomSalinity (km)`, Wind_speed=`Wind Speed (m/s)`)%>%
  mutate(Date=dmy(Date))%>%
  arrange(Date)%>%
  mutate(data=if_else(Date>test_train_cutoff, "Testing", "Training"),
         NDOI_change=NDOI-lag(NDOI),
         NDOI_change_mag=abs(NDOI_change),
         NDOI_change_sign=sign(NDOI_change),
         NDOI_prop_change=NDOI_change/lag(NDOI),
         Weak_ebb_flood_1=lag(Weak_ebb_flood),
         Weak_ebb_flood_2=lag(Weak_ebb_flood, 2))

d_train<-d%>%
  filter(data=="Training")

# Create variable with residuals of X2_surface by NDOI
X2_surface_model<-lm(X2_surface~log(NDOI), data=d_train)

d_train<-d_train%>%
  mutate(X2_surface_resid=residuals(X2_surface_model))

d_long<-d_train%>%
  select(-NDOI_change_mag, -NDOI_change_sign)%>%
  pivot_longer(cols = c(NDOI, NDOI_change, NDOI_prop_change, 
                        Weak_ebb_flood, Stratification, Wind_speed,
                        X2_surface, X2_bottom),
               names_to="variable",
               values_to="value")
```

# Plots

full timeseries
```{r, fig.height=10}
ggplot(d_long, aes(x=Date, y=value, color=variable))+
  geom_line()+
  facet_wrap(~variable, ncol=1, scales="free_y")+
  scale_color_viridis_d()+
  theme_bw()+
  theme(legend.position="none")
```

zoomed-in timeseries for May through September of 2022
```{r, fig.height=10}
ggplot(filter(d_long, year(Date)==2022 & month(Date)>=5), aes(x=Date, y=value, color=variable))+
  geom_line()+
  facet_wrap(~variable, ncol=1, scales="free_y")+
  scale_color_viridis_d()+
  theme_bw()+
  theme(legend.position="none")
```

Weak_ebb_flood vs stratification
```{r}
ggplot(d_train, aes(x=Weak_ebb_flood, y=Stratification, color=X2_surface))+
  geom_point()+
  scale_color_viridis_c()+
  theme_bw()
```

X2 vs stratification
```{r}
ggplot(d_train, aes(x=X2_surface, y=Stratification, color=Weak_ebb_flood))+
  geom_point()+
  scale_color_viridis_c()+
  theme_bw()
```

X2 residual vs stratification
```{r}
ggplot(d_train, aes(x=X2_surface_resid, y=Stratification, color=Weak_ebb_flood))+
  geom_point()+
  scale_color_viridis_c()+
  theme_bw()
```

Wind speed vs stratification
```{r}
ggplot(d_train, aes(x=Wind_speed, y=Stratification, color=log(NDOI)))+
  geom_point()+
  scale_color_viridis_c()+
  theme_bw()
```

NDOI vs stratification
```{r}
ggplot(d_train, aes(x=NDOI, y=Stratification, color=Weak_ebb_flood))+
  geom_point()+
  scale_x_log10()+
  scale_color_viridis_c()+
  theme_bw()
```

NDOI change vs stratification
```{r}
ggplot(d_train, aes(x=abs(NDOI_change_mag), y=Stratification, color=factor(NDOI_change_sign)))+
  geom_point()+
  scale_x_log10()+
  scale_color_viridis_d()+
  theme_bw()
```

NDOI prop change vs stratification
```{r}
ggplot(d_train, aes(x=abs(NDOI_prop_change), y=Stratification, color=Weak_ebb_flood))+
  geom_point()+
  #scale_x_log10()+
  scale_color_viridis_c()+
  theme_bw()
```

NDOI vs X2 with trend line used for residuals
```{r}
ggplot(d_train, aes(x=log(NDOI), y=X2_surface, color=Weak_ebb_flood))+
  geom_point()+
  geom_abline(slope=coef(X2_surface_model)[2], intercept=coef(X2_surface_model)[1], color="red")+
  scale_color_viridis_c()+
  theme_bw()
```

NDOI change vs X2
```{r}
ggplot(d_train, aes(x=NDOI_change_mag, y=X2_surface, color=log(NDOI)))+
  geom_point()+
  scale_x_log10()+
  scale_color_viridis_c()+
  theme_bw()
```

Pairs plot of all variables
```{r, warning=FALSE, fig.height=10, fig.width=10}
ggpairs(d_train)+
  theme_bw()
```

Takeaways to inform model construction
1. NDOI_ _change_mag is highly correlated with NDOI, but NDOI_change is not
2. X2 is highly correlated with NDOI
3. Weak_ebb_flood not correlated with anything (except slightly with stratification)

# Model

Prepare data
```{r}
d_model<-d_train%>%
  drop_na()%>%
  mutate(across(c(Stratification, NDOI), list(l=~log(.x))),
         across(c(NDOI, NDOI_l, NDOI_change, Weak_ebb_flood, Weak_ebb_flood_1, Weak_ebb_flood_2, Wind_speed, X2_surface_resid), list(s=~(.x-mean(.x))/sd(.x))))
```

model function
```{r}
gam_model<-function(form, model_data=d_model, model_family="gaussian", plot_acf=FALSE){
  par(mfrow=c(2,2))
  NOAR<-bam(form, data=model_data, family=model_family) # first fit model with autoregressive (AR) term
  par(mfrow=c(1,1))
  r <- start_value_rho(NOAR, plot=plot_acf) # Estimate rho for autoregressive term
  if(model_family=="gaussian"){
    m<-bam(form, data=model_data, rho=r, family=model_family) # fit autoregressive model
  }else{
    m<-bam(form, data=model_data, rho=r, family=model_family, discrete=T) # fit autoregressive model
  }
  if(plot_acf){
    suppressWarnings(acf_resid(m))
  }
  par(mfrow=c(2,2))
  gam.check(m)
  print(summary(m))
  par(mfrow=c(1,1))
  return(m)
}
```


## m1 

```{r, warning=FALSE}
m1<-gam_model(Stratification_l ~ s(NDOI_l_s) +s(NDOI_change_s) + s(Weak_ebb_flood_s) + s(Wind_speed_s) + s(X2_surface_resid_s), plot_acf=TRUE, model_family="scat")
```

Only `X2_surface_resid_s` is showing as a nonlinear effect, so making the rest linear

## m2

```{r, warning=FALSE}
m2<-gam_model(Stratification_l ~ NDOI_l_s + NDOI_change_s + Weak_ebb_flood_s + Wind_speed_s + s(X2_surface_resid_s), model_family="scat")
AIC(m1, m2)
```

`NDOI_change_s` doesn't have a significant effect, so removing it

## m3

```{r, warning=FALSE}
m3<-gam_model(Stratification_l ~ NDOI_l_s  + Weak_ebb_flood_s + Wind_speed_s + s(X2_surface_resid_s), model_family="scat")
AIC(m1, m2, m3)
```

Nope, it was better with `NDOI_change_s` 

Do we need interactions?

## m4

```{r, warning=FALSE}
m4<-gam_model(Stratification_l ~ (NDOI_l_s + Wind_speed_s + NDOI_change_s + Weak_ebb_flood_s)^2 + s(X2_surface_resid_s), model_family="scat")
AIC(m1, m2, m3, m4)
```

Only NDOI_l_s:Wind_speed_s is significant

## m5

```{r, warning=FALSE}
m5<-gam_model(Stratification_l ~ NDOI_l_s * Wind_speed_s + NDOI_change_s + Weak_ebb_flood_s + s(X2_surface_resid_s), model_family="scat")
AIC(m1, m2, m3, m4, m5)
```

Model is better with NDOI_l_s:Wind_speed_s interaction so moving forward with m5

# Plot m5 results

Generate predictions
```{r}
newdata<-expand_grid(NDOI=seq(min(d_model$NDOI), max(d_model$NDOI), length.out=10),
                     Weak_ebb_flood=seq(min(d_model$Weak_ebb_flood), max(d_model$Weak_ebb_flood), length.out=10),
                     Wind_speed=seq(min(d_model$Wind_speed), max(d_model$Wind_speed), length.out=10),
                     NDOI_change=seq(min(d_model$NDOI_change), max(d_model$NDOI_change), length.out=10),
                     X2_surface_resid=seq(min(d_model$X2_surface_resid), max(d_model$X2_surface_resid), length.out=10))%>%
  mutate(NDOI_l=log(NDOI),
         NDOI_l_s=(NDOI_l-mean(d_model$NDOI_l))/sd(d_model$NDOI_l),
         Weak_ebb_flood_s=(Weak_ebb_flood-mean(d_model$Weak_ebb_flood))/sd(d_model$Weak_ebb_flood),
         Wind_speed_s=(Wind_speed-mean(d_model$Wind_speed))/sd(d_model$Wind_speed),
         NDOI_change_s=(NDOI_change-mean(d_model$NDOI_change))/sd(d_model$NDOI_change),
         X2_surface_resid_s=(X2_surface_resid-mean(d_model$X2_surface_resid))/sd(d_model$X2_surface_resid))

m4b_pred<-predict(m5, newdata=newdata, se=TRUE)

newdata_pred<-newdata%>%
  mutate(Stratification_l=m4b_pred$fit,
         Stratification_l_se=m4b_pred$se.fit,
         Stratification_l_u95=Stratification_l+Stratification_l_se*qnorm(0.95),
         Stratification_l_l95=Stratification_l-Stratification_l_se*qnorm(0.95),
         across(c(Stratification_l, Stratification_l_u95, Stratification_l_l95), exp))%>%
  rename_with(~str_remove(.x, fixed("_l")), c(Stratification_l, Stratification_l_u95, Stratification_l_l95))
```

Plotting function
```{r}
strat_pred_plot<-function(xvar, colorvar){
  othervars<-c("NDOI", "Weak_ebb_flood", "NDOI_change", "Wind_speed", "X2_surface_resid")
  othervars<-othervars[-which(othervars%in%c(xvar, colorvar))]
  
  ggplot(
    filter(newdata_pred, 
           .data[[colorvar]]%in%unique(.data[[colorvar]])[c(1, 5, 10)] &
             .data[[othervars[1]]]%in%unique(.data[[othervars[1]]])[5] &
             .data[[othervars[2]]]%in%unique(.data[[othervars[2]]])[5] &
             .data[[othervars[3]]]%in%unique(.data[[othervars[3]]])[5]), aes(x=.data[[xvar]], y=Stratification, ymin=Stratification_l95, ymax=Stratification_u95, fill=.data[[colorvar]], color=.data[[colorvar]], group=.data[[colorvar]]))+
    geom_ribbon(alpha=0.4, color=NA)+
    geom_line()+
    scale_fill_viridis_c(aesthetics = c("color", "fill"))+
    theme_bw()
}
```


plot by NDOI
```{r}
strat_pred_plot(xvar="NDOI", colorvar="Weak_ebb_flood")
```

plot by Weak_ebb_flood
```{r}
strat_pred_plot(xvar="Weak_ebb_flood", colorvar="NDOI")
```

plot by Wind speed
```{r}
strat_pred_plot(xvar="Wind_speed", colorvar="NDOI")
```

plot by NDOI change
```{r}
strat_pred_plot(xvar="NDOI_change", colorvar="NDOI")
```

plot by NDOI change
```{r}
strat_pred_plot(xvar="X2_surface_resid", colorvar="NDOI")
```

# Cross validation

Testing data
```{r}
d_test<-d%>%
  filter(data=="Testing" & !is.na(X2_surface))


# Create variable with residuals of X2_surface by NDOI
X2_surface_model_train<-lm(X2_surface~log(NDOI), data=d_test)

d_test<-d_test%>%
  mutate(X2_surface_resid=residuals(X2_surface_model_train))%>%
  drop_na()%>%
  mutate(across(c(Stratification, NDOI), list(l=~log(.x))),
         across(c(NDOI, NDOI_l, NDOI_change, Weak_ebb_flood, Weak_ebb_flood_1, Weak_ebb_flood_2, Wind_speed, X2_surface_resid), list(s=~(.x-mean(d_model[[cur_column()]]))/sd(d_model[[cur_column()]]))))

d_test_pred<-d_test%>% # center and scale data based on original model inputs
  mutate(stratification_pred_l=predict(m5, newdata=.), # Generate model predictions
         stratification_pred=exp(stratification_pred_l),
         Stratification_diff=stratification_pred-Stratification,
         Stratification_diff_prop=Stratification_diff/Stratification)
```

Cross-validation statistics
```{r}
# RMSE
sqrt(mean((d_test_pred$Stratification_diff)^2))

# Mean difference (for bias)
mean(d_test_pred$Stratification_diff)

# Pearson's correlation
cor(d_test_pred$Stratification, d_test_pred$stratification_pred, method = 'pearson')
```


plot measured vs predicted
```{r}
ggplot(d_test_pred, aes(x=Stratification, y=stratification_pred))+
  geom_point()+
  geom_abline(intercept = 0, slope=1, color="red")+
  theme_bw()
```

plot diff vs outflow
```{r}
ggplot(d_test_pred, aes(x=NDOI, y=Stratification_diff))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  theme_bw()
```

plot diff vs outflow change
```{r}
ggplot(d_test_pred, aes(x=NDOI_change, y=Stratification_diff))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  theme_bw()
```

plot diff vs Weak_ebb_flood
```{r}
ggplot(d_test_pred, aes(x=Weak_ebb_flood, y=Stratification_diff))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  theme_bw()
```

plot diff vs Wind_speed
```{r}
ggplot(d_test_pred, aes(x=Wind_speed, y=Stratification_diff))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  theme_bw()
```

plot diff vs X2_surface_resid
```{r}
ggplot(d_test_pred, aes(x=X2_surface_resid, y=Stratification_diff))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  theme_bw()
```

plot diff vs X2_surface
```{r}
ggplot(d_test_pred, aes(x=X2_surface, y=Stratification_diff))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  theme_bw()
```

# Predict

Plot measured and model-predicted stratification
```{r}
d_model_pred<-d_model%>%
  bind_rows(d_test)%>%
  mutate(stratification_pred_l=predict(m5, newdata=.),
         stratification_pred=exp(stratification_pred_l))

d_model_pred_long<-d_model_pred%>%
  select(Date, Stratification, stratification_pred)%>%
  pivot_longer(cols = c(stratification_pred, Stratification), names_to = "variable", values_to="stratification")%>%
  mutate(variable=recode(variable, stratification_pred="modeled", Stratification="measured"))%>%
  complete(Date=seq.Date(min(d$Date), max(d$Date), by = "day"), variable)

ggplot(d_model_pred_long, aes(x=Date, y=stratification, color=variable))+
  geom_line()+
  geom_vline(xintercept=test_train_cutoff, color="red", linetype=2)+
  scale_color_viridis_d(end=0.8)+
  theme_bw()
```

Plot measured vs modeled stratification
```{r}
ggplot(d_model_pred, aes(x=Stratification, y=stratification_pred, color=data))+
  geom_abline(intercept=0, slope=1, color="red")+
  geom_point(alpha=0.5)+
  scale_color_viridis_d(end=0.8)+
  xlab("Measured stratification")+
  ylab("Modeled stratification")+
  theme_bw()
```

# Download modeled stratification

```{r, echo=FALSE}
download_this(
  .data = d_model_pred,
  output_name = paste("Modeled stratification data", Sys.Date()),
  output_extension = ".csv",
  button_label = "Download data",
  button_type = "warning",
  has_icon = TRUE,
  icon = "fa fa-save",
  csv2 = FALSE
)
```
