---
title: "stratification-model"
author: "Sam Bashevkin"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = 'docs',
      knit_root_dir = "../",
      envir = globalenv()
    )
    })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

packages
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(GGally)
library(mgcv)
library(itsadug)
```

read data
```{r}
d<-read_csv("data/X2_ModelInputs_v1.csv")%>%
  filter(if_any(everything(), ~ !is.na(.x)))%>% # Remove empty rows
  rename(NDOI=`NDOI (cfs)`, Diurnal_inequality=`Diurnal Inequality (m)`, Stratification=`Stratification (PSU)`,
         X2_surface=`X2 Top Salinity (km)`, X2_bottom=`X2 Bottom Salinity (km)`)%>%
  mutate(Date=dmy(Date))%>%
  arrange(Date)%>%
  mutate(NDOI_change=NDOI-lag(NDOI),
         NDOI_change_mag=abs(NDOI_change),
         NDOI_change_sign=sign(NDOI_change),
         NDOI_prop_change=NDOI_change/lag(NDOI)) 

d_long<-d%>%
  select(-NDOI_change_mag, -NDOI_change_sign)%>%
  pivot_longer(cols = c(NDOI, NDOI_change, NDOI_prop_change, 
                        Diurnal_inequality, Stratification,
                        X2_surface, X2_bottom),
               names_to="variable",
               values_to="value")
```

# Plots

full timeseries
```{r, fig.height=10}
ggplot(d_long, aes(x=Date, y=value, color=variable))+
  geom_line()+
  facet_wrap(~variable, ncol=1, scales="free_y")+
  scale_color_viridis_d()+
  theme_bw()+
  theme(legend.position="none")
```

zoomed-in timeseries for May through September of 2022
```{r, fig.height=10}
ggplot(filter(d_long, year(Date)==2022 & month(Date)>=5), aes(x=Date, y=value, color=variable))+
  geom_line()+
  facet_wrap(~variable, ncol=1, scales="free_y")+
  scale_color_viridis_d()+
  theme_bw()+
  theme(legend.position="none")
```

Diurnal_inequality vs stratification
```{r}
ggplot(d, aes(x=Diurnal_inequality, y=Stratification, color=X2_surface))+
  geom_point()+
  scale_color_viridis_c()+
  theme_bw()
```

X2 vs stratification
```{r}
ggplot(d, aes(x=X2_surface, y=Stratification, color=Diurnal_inequality))+
  geom_point()+
  scale_color_viridis_c()+
  theme_bw()
```

NDOI vs stratification
```{r}
ggplot(d, aes(x=NDOI, y=Stratification, color=Diurnal_inequality))+
  geom_point()+
  scale_x_log10()+
  scale_color_viridis_c()+
  theme_bw()
```

NDOI change vs stratification
```{r}
ggplot(d, aes(x=abs(NDOI_change_mag), y=Stratification, color=factor(NDOI_change_sign)))+
  geom_point()+
  scale_x_log10()+
  scale_color_viridis_d()+
  theme_bw()
```

NDOI prop change vs stratification
```{r}
ggplot(d, aes(x=abs(NDOI_prop_change), y=Stratification, color=Diurnal_inequality))+
  geom_point()+
  #scale_x_log10()+
  scale_color_viridis_c()+
  theme_bw()
```

NDOI vs X2
```{r}
ggplot(d, aes(x=NDOI, y=X2_surface, color=Diurnal_inequality))+
  geom_point()+
  scale_x_log10()+
  scale_color_viridis_c()+
  theme_bw()
```

NDOI change vs X2
```{r}
ggplot(d, aes(x=NDOI_change_mag, y=X2_surface, color=log(NDOI)))+
  geom_point()+
  scale_x_log10()+
  scale_color_viridis_c()+
  theme_bw()
```

Pairs plot of all variables
```{r, warning=FALSE, fig.height=10, fig.width=10}
ggpairs(d)+
  theme_bw()
```

Takeaways to inform model construction
1. NDOI_ _change_mag is highly correlated with NDOI, but NDOI_change is not
2. X2 is highly correlated with NDOI
3. Diurnal_inequality not correlated with anything (except slightly with stratification)

# Model

Prepare data
```{r}
d_model<-d%>%
  drop_na()%>%
  mutate(across(c(Stratification, NDOI), list(l=~log(.x))),
         across(c(NDOI, NDOI_l, NDOI_change, Diurnal_inequality), list(s=~(.x-mean(.x))/sd(.x))))
```

## m1 

Stratification_l ~ te(NDOI_s, NDOI_change_s, Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m1_NOAR<-bam(Stratification_l ~ te(NDOI_s, NDOI_change_s, Diurnal_inequality_s), data=d_model) # first fit model with autoregressive (AR) term
par(mfrow=c(1,1))
m1_r <- start_value_rho(m1_NOAR, plot=TRUE) # Estimate rho for autoregressive term
m1<-bam(Stratification_l ~ te(NDOI_s, NDOI_change_s, Diurnal_inequality_s), data=d_model, rho=m1_r) # fit autoregressive model
par(mfrow=c(2,2))
gam.check(m1)
summary(m1)
```

## m2

Stratification_l ~ s(NDOI_s) +s(NDOI_change_s) + s(Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m2_NOAR<-bam(Stratification_l ~ s(NDOI_s) +s(NDOI_change_s) + s(Diurnal_inequality_s), data=d_model)
m2_r <- start_value_rho(m2_NOAR, plot=FALSE)
m2<-bam(Stratification_l ~ s(NDOI_s) +s(NDOI_change_s) + s(Diurnal_inequality_s), data=d_model, rho=m2_r)
gam.check(m2)
summary(m2)

AIC(m1, m2)
```

NDOI_change_s has an edf of 1 indicating the relationship is linear, so changing to linear

## m3

Stratification_l ~ s(NDOI_s) + NDOI_change_s + s(Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m3_NOAR<-bam(Stratification_l ~ s(NDOI_s) + NDOI_change_s + s(Diurnal_inequality_s), data=d_model)
m3_r <- start_value_rho(m3_NOAR, plot=FALSE)
m3<-bam(Stratification_l ~ s(NDOI_s) + NDOI_change_s + s(Diurnal_inequality_s), data=d_model, rho=m3_r)
gam.check(m3)
summary(m3)

AIC(m1, m2, m3)
```

Same model as above, it must have just automatically converted to linear when it detected it was such. NDOI_change_s has no sig effect so dropping it. 

## m4

Stratification_l ~ s(NDOI_s) + s(Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m4_NOAR<-bam(Stratification_l ~ s(NDOI_s) + s(Diurnal_inequality_s), data=d_model)
m4_r <- start_value_rho(m4_NOAR, plot=FALSE)
m4<-bam(Stratification_l ~ s(NDOI_s) + s(Diurnal_inequality_s), data=d_model, rho=m4_r)
gam.check(m4)
summary(m4)

AIC(m1, m2, m4)
```

Slightly better and more parsimonious without NDOI_change, so trying an interaction between 2 remaining variables

## m5

Stratification_l ~ te(NDOI_s, Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m5_NOAR<-bam(Stratification_l ~ te(NDOI_s, Diurnal_inequality_s), data=d_model)
m5_r <- start_value_rho(m5_NOAR, plot=FALSE)
m5<-bam(Stratification_l ~ te(NDOI_s, Diurnal_inequality_s), data=d_model, rho=m5_r)
gam.check(m5)
summary(m5)

AIC(m1, m2, m4, m5)
```

This is worse, so m4 is best so far

Now trying with NDOI log transformed

## m1b

Stratification_l ~ te(NDOI_l_s, NDOI_change_s, Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m1b_NOAR<-bam(Stratification_l ~ te(NDOI_l_s, NDOI_change_s, Diurnal_inequality_s), data=d_model)
m1b_r <- start_value_rho(m1b_NOAR, plot=FALSE)
m1b<-bam(Stratification_l ~ te(NDOI_l_s, NDOI_change_s, Diurnal_inequality_s), data=d_model, rho=m1b_r)
gam.check(m1b)
summary(m1b)
```

## m2b

Stratification_l ~ s(NDOI_l_s) +s(NDOI_change_s) + s(Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m2b_NOAR<-bam(Stratification_l ~ s(NDOI_l_s) +s(NDOI_change_s) + s(Diurnal_inequality_s), data=d_model)
m2b_r <- start_value_rho(m2b_NOAR, plot=FALSE)
m2b<-bam(Stratification_l ~ s(NDOI_l_s) +s(NDOI_change_s) + s(Diurnal_inequality_s), data=d_model, rho=m2b_r)
gam.check(m2b)
summary(m2b)

AIC(m1b, m2b)
```

Slightly better and more parsimonious. Only Diurnal_frequency is nonlinear, so using linear versions of the other 2

## m3b

Stratification_l ~ NDOI_l_s + NDOI_change_s + s(Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m3b_NOAR<-bam(Stratification_l ~ NDOI_l_s + NDOI_change_s + s(Diurnal_inequality_s), data=d_model)
m3b_r <- start_value_rho(m3b_NOAR, plot=FALSE)
m3b<-bam(Stratification_l ~ NDOI_l_s + NDOI_change_s + s(Diurnal_inequality_s), data=d_model, rho=m3b_r)
gam.check(m3b)
summary(m3b)

AIC(m1b, m2b, m3b)
```

Better, but NDOI_change isn't significant, so dropping it

## m4b

Stratification_l ~ NDOI_l_s + s(Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m4b_NOAR<-bam(Stratification_l ~ NDOI_l_s + s(Diurnal_inequality_s), data=d_model)
m4b_r <- start_value_rho(m4b_NOAR, plot=FALSE)
m4b<-bam(Stratification_l ~ NDOI_l_s + s(Diurnal_inequality_s), data=d_model, rho=m4b_r)
gam.check(m4b)
summary(m4b)

AIC(m1b, m2b, m3b, m4b)
```

Basically the same and more parsimonious. Now trying an interaction among 2 remaining variables. 

## m5b

Stratification_l ~ te(NDOI_l_s, Diurnal_inequality_s)
```{r}
par(mfrow=c(2,2))
m5b_NOAR<-bam(Stratification_l ~ te(NDOI_l_s, Diurnal_inequality_s), data=d_model)
m5b_r <- start_value_rho(m5b_NOAR, plot=FALSE)
m5b<-bam(Stratification_l ~ te(NDOI_l_s, Diurnal_inequality_s), data=d_model, rho=m5b_r)
gam.check(m5b)
summary(m5b)

AIC(m1b, m2b, m3b, m4b, m5b)
```

Worse, so m4b is best of this round, now comparing with m4
```{r}
AIC(m4, m4b)
```

m4b is best, moving forward with that

# Plot m4b results

Generate predictions
```{r}
newdata<-expand_grid(NDOI=seq(min(d_model$NDOI), max(d_model$NDOI), length.out=50),
                     Diurnal_inequality=seq(min(d_model$Diurnal_inequality), max(d_model$Diurnal_inequality), length.out=50))%>%
  mutate(NDOI_l=log(NDOI),
         NDOI_l_s=(NDOI_l-mean(d_model$NDOI_l))/sd(d_model$NDOI_l),
         Diurnal_inequality_s=(Diurnal_inequality-mean(d_model$Diurnal_inequality))/sd(d_model$Diurnal_inequality))

m4b_pred<-predict(m4b, newdata=newdata, se=TRUE)

newdata_pred<-newdata%>%
  mutate(Stratification_l=m4b_pred$fit,
         Stratification_l_se=m4b_pred$se.fit,
         Stratification_l_u95=Stratification_l+Stratification_l_se*qnorm(0.95),
         Stratification_l_l95=Stratification_l-Stratification_l_se*qnorm(0.95),
         across(c(Stratification_l, Stratification_l_u95, Stratification_l_l95), exp))%>%
  rename_with(~str_remove(.x, fixed("_l")), c(Stratification_l, Stratification_l_u95, Stratification_l_l95))
```


plot by NDOI
```{r}
ggplot(filter(newdata_pred, Diurnal_inequality%in%unique(Diurnal_inequality)[c(1, 25, 50)]), aes(x=NDOI, y=Stratification, ymin=Stratification_l95, ymax=Stratification_u95, fill=Diurnal_inequality, color=Diurnal_inequality, group=Diurnal_inequality))+
  geom_ribbon(alpha=0.4, color=NA)+
  geom_line()+
  scale_fill_viridis_c(aesthetics = c("color", "fill"))+
  theme_bw()
```


plot by Diurnal_inequality
```{r}
ggplot(filter(newdata_pred, NDOI%in%unique(NDOI)[c(1, 25, 50)]), aes(x=Diurnal_inequality, y=Stratification, ymin=Stratification_l95, ymax=Stratification_u95, fill=NDOI, color=NDOI, group=NDOI))+
  geom_ribbon(alpha=0.4, color=NA)+
  geom_line()+
  scale_fill_viridis_c(aesthetics = c("color", "fill"))+
  theme_bw()
```
